---
title:  "GSE_Analysis"
output: html_notebook
name:   Michael McGrath
uni:    mwm2166
---

# Preform a gene set enrichment analysis on the genes based on the differential expression between the three groups to investigate the pathways driving the patterns found in the DE analysis (see DE_Analysis.Rmd).

## Install packages.

```{r}
# install required CRAN packages
for (pkg in c("BiocManager", "data.table", "httr", "FactoMineR", "statmod")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# install required Bioc packages
for (pkg in c("limma", "edgeR", "GO.db", "org.Hs.eg.db")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, update = FALSE, ask = FALSE)
  }
}
```

## Need to convert the gene symbols to gene ids for the gene enrichment.

```{r}
library(httr)
library(data.table)

GENE_SYM_ID_LINK_BASE <- "https://www.genenames.org/cgi-bin/download/custom?col=gd_app_sym&col=gd_pub_eg_id&status=Approved&hgnc_dbtag=on&order_by=gd_app_sym_sort&format=text&submit=submit"
response <- GET(GENE_SYM_ID_LINK_BASE)
gene_sym_id <- data.frame(fread(text=httr::content(response, "parsed"), header=TRUE))
colnames(gene_sym_id) <- c("gene_symbol","gene_id")

gene_sym_id <- gene_sym_id[apply(gene_sym_id == "", 1, sum) == 0,]
gene_sym_id <- gene_sym_id[apply(is.na(gene_sym_id), 1, sum) == 0,]

gene_sym_id <- gene_sym_id[!duplicated(gene_sym_id$gene_id), ]
rownames(gene_sym_id) <- gene_sym_id$gene_symbol
```

## Load in the design matrix and fit DGEGLM object, which was output in by the QLFit method in the DE analysis (see DE_Analysis.Rmd).

```{r}
load("data/fit.RData")
load("data/design.RData")
```

## Preform the gene set enrichment analysis on each of the three contrasts.

```{r}
library(limma)
library(edgeR)
n <- 20
contrast <- makeContrasts(SC2-no_virus, 
                           levels = design)
qlf <- glmQLFTest(fit, contrast=contrast)
go_res <- goana(qlf, geneid=gene_sym_id[as.vector(unlist(qlf$genes)), "gene_id"])
top <- topGO(go_res, number=n)[, c('Term', 'Ont', 'P.Up', 'P.Down')]
top['PValue']    <- apply(top[, c('P.Up', 'P.Down')], 1, min)
top['regulated'] <- apply(top[, c('P.Up', 'P.Down')], 1,
                          function(x){ifelse(x[1] < x[2],
                                             "up-regulated",
                                             "down-regulated")})
top_SC2_vs_no_virus <- top[, c('Term', 'Ont', 'PValue', 'regulated')]
top_SC2_vs_no_virus
```

```{r}
n <- 50
contrast <- makeContrasts(SC2-other_virus, 
                           levels = design)
qlf <- glmQLFTest(fit, contrast=contrast)
SC2_vs_other_virus <- as.data.frame(topTags(qlf, n=Inf))
go_res <- goana(qlf, geneid=gene_sym_id[as.vector(unlist(qlf$genes)), "gene_id"])
top <- topGO(go_res, number=n)[, c('Term', 'Ont', 'P.Up', 'P.Down')]
top['PValue']    <- apply(top[, c('P.Up', 'P.Down')], 1, min)
top['regulated'] <- apply(top[, c('P.Up', 'P.Down')], 1,
                          function(x){ifelse(x[1] < x[2],
                                             "up-regulated",
                                             "down-regulated")})
top_SC2_vs_other_virus <- top[, c('Term', 'Ont', 'PValue', 'regulated')]
top_SC2_vs_other_virus
```

```{r}
n <- 30
contrast <- makeContrasts(other_virus-no_virus, 
                           levels = design)
qlf <- glmQLFTest(fit, contrast=contrast)
other_virus_vs_no_virus <- as.data.frame(topTags(qlf, n=Inf))
go_res <- goana(qlf, geneid=gene_sym_id[as.vector(unlist(qlf$genes)), "gene_id"])
top <- topGO(go_res, number=n)[, c('Term', 'Ont', 'P.Up', 'P.Down')]
top['PValue']    <- apply(top[, c('P.Up', 'P.Down')], 1, min)
top['regulated'] <- apply(top[, c('P.Up', 'P.Down')], 1,
                          function(x){ifelse(x[1] < x[2],
                                             "up-regulated",
                                             "down-regulated")})
top_other_virus_vs_no_virus <- top[, c('Term', 'Ont', 'PValue', 'regulated')]
top_other_virus_vs_no_virus
```

```{r}
library(GO.db)
top_SC2_vs_other_virus["Definition"] <- Definition(row.names(top_SC2_vs_other_virus))
```

```{r}
ind = 0
for (i in row.names(top_SC2_vs_other_virus)[1:20]) {
  ind <- ind + 1
  print(cat(c(toString(ind), ":",
              toString(top_SC2_vs_other_virus[i, "Term"]), ":", "\n   ",
              toString(top_SC2_vs_other_virus[i, "Definition"]), "(",
              toString(top_SC2_vs_other_virus[i, "regulated"]), ")"
              )
            )
        )
}
```
